---
title: "syllabus"
format: pdf
editor: visual
---

## Syllabus: Introduction to Social Data Science

------------------------------------------------------------------------

#### **Course Information**

-   **Course Code:** ULI2008.1 – Sosyal Veri Bilimine Giriş / PSIR3066 – Introduction to Data Science

-   **Schedule:** Monday, 3 hours (14:00–17:00)

-   **Classroom:** Faculty of Political Science, Room RTE.S1.124

-   **Office Hours:** Monday, 09:00–11:00

-   **Office Location:** Faculty of Political Science, Room RTE.S1.131

-   **Credits & ECTS:** 3 credits, ECTS: 5

-   **Course Type:** Elective

-   **Instructor:** Assoc. Prof. Hakan Mehmetcik

-   **Contact:** hakan.mehmetcik\@marmara.edu.tr

------------------------------------------------------------------------

### **Course Objectives** 

This course provides an introduction to **R**, a free and open-source statistical computing environment and programming language widely used in **data analysis, statistical research, and industrial applications**. The primary goal of this course is to equip students with the ability to use **R for problem-solving and practical applications** in social data science. Students will gain hands-on experience with data manipulation, visualization, statistical computation, and analysis beyond basic spreadsheet tools such as Excel.

R is a versatile programming language that enables efficient **data processing, statistical modeling, and visualization**. It runs on multiple platforms, including **UNIX, Linux, Windows, and macOS**.

This course is designed for students who want to learn the fundamentals of R and its applications in social data science. The content focuses on:

-   The **basics of the R environment** and how to navigate it effectively.

-   **Arithmetic and logical operators**, data structures, variables, and data types in R.

-   **Data manipulation and transformation** techniques, including working with R packages for visualization, reporting, and statistical analysis.

-   **Importing, processing, and transforming datasets** for analytical purposes.

-   Core **data science concepts**, including **parametric and non-parametric statistical tests**.

By the end of the course, students will have a strong foundation in **R programming** and its **applications in data science**, enabling them to analyze and visualize real-world datasets effectively.

### **Course Learning Outcomes**

By the end of this course, students will be able to:

-   **Use R and RStudio effectively** for data analysis and programming.

-   **Process and analyze data** using R’s built-in functions and libraries.

-   **Create visualizations and reports** to communicate data-driven insights.

-   **Apply statistical methods** for data exploration and hypothesis testing.

These skills will provide students with a strong foundation in **data analysis and statistical computing**, preparing them for **more advanced data science and problem-solving tasks**.

------------------------------------------------------------------------

## **Course Readings**

#### **Primary Textbooks (Required Readings)**

[Modern Data Science with R](https://mdsr-book.github.io/mdsr3e/)

[R for Data Science](https://r4ds.had.co.nz/)

### Supplementary Resources (Optional Readings & References)

-   [An Introduction to R](http://cran.r-project.org/doc/manuals/R-intro.html) ([pdf](http://cran.r-project.org/doc/manuals/R-intro.pdf)) by W. N. Venables, D. M. Smith, and the R Development Core Team
-   [Advanced R (2nd Ed)](https://adv-r.hadley.nz/) by Hadley Wickham (advanced)
-   [Intro to R video lectures](https://www.youtube.com/playlist?list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP) by Google Developers
-   [R Programming wikibook](http://en.wikibooks.org/wiki/R_Programming)
-   [Using R for Data Analysis and Graphics](http://cran.r-project.org/doc/contrib/usingR.pdf) by J. H. Maindonald
-   [The R Inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf) by Patrick Burns (advanced)

### **Required Materials for the Course**

#### **Hardware Requirement**

-   Students **must bring a laptop** to class for hands-on exercises.

#### **Software Requirement**

To participate in this course, students need to install **R and RStudio** on their personal computers.

1.  **Download and Install R**

    -   First, download and install **R** from [The R Project](https://cloud.r-project.org/).

2.  **Download and Install RStudio**

    -   After installing R, download and install **RStudio** from [Posit](https://www.posit.co/).

3.  **Install Required R Packages**

    -   During the course, students will need to install and use the **tidyverse** and **RMarkdown** packages. Instructions for installing these and other required packages will be provided in class.

------------------------------------------------------------------------

### **Getting Help**

If you need assistance, consider the following steps:

1.  **Self-Help Resources**

    -   Use **R Help (?command)** within RStudio.

    -   Search for solutions on **Google** or **AI platforms** (e.g., **ChatGPT, DeepSeek**).

2.  **Ask Questions in Google Classroom**

    -   Post your questions in the **Google Classroom discussion board** for peer and instructor support.

    -   If you know the answer to a classmate’s question, feel free to help!

3.  **Schedule a Virtual Meeting**

    -   If additional support is needed, students can **schedule a virtual meeting** via **Google Meet** with the instructor.

------------------------------------------------------------------------

### **Attendance and Participation Requirements**

Research indicates that **regular attendance** is one of the strongest predictors of academic success.

According to the university’s **Education and Teaching Regulations**, students are required to:

-   Attend at least **70% of theoretical classes**.

-   Attend at least **80% of practical sessions** (if applicable).

🔹 **Please consider these attendance requirements carefully before enrolling in this course!**

------------------------------------------------------------------------

### **Grading & Final Project Requirements**

#### **Grading Breakdown**

-   **Midterm Project – Phase 1:** 40%

-   **Final Project – Phase 2:** 50%

-   **Participation:** 10%

📢 **For exact deadlines, please check the university’s online system (OBYS).**

------------------------------------------------------------------------

Each student is required to conduct a **data analysis using R** based on a dataset sourced from [**Data is Plural**](https://www.data-is-plural.com/archive/). The project consists of two phases:

#### **Midterm Project – Phase 1:** *(Problem Definition & Data Collection)*

1.  **Problem Definition & Data Selection:**

    -   Identify a **social issue** you want to analyze.

    -   Formulate a **clear research question** and determine the relevant dataset.

    -   Data **must be selected from** [Data is Plural’s dataset archive](https://www.data-is-plural.com/archive/).

2.  **Data Cleaning & Exploratory Data Analysis (EDA):**

    -   Prepare the dataset for analysis by handling **missing values, outliers, and irrelevant data**.

    -   Conduct **Exploratory Data Analysis (EDA)** to identify patterns and relationships in the data.

------------------------------------------------------------------------

#### **Final Project – Phase 2:** *(Data Modeling & Reporting)*

3.  **Data Modeling:**

    -   Based on EDA insights, select **appropriate statistical or machine learning models** to answer your research question.

    -   Justify your model choice and interpret the results.

4.  **Presentation & Report Writing:**

    -   Clearly present your findings through **visualizations, a written report, or a presentation**.

    -   The report should include all project phases, from **problem definition to conclusions**.

📢 **For exact submission deadlines, please check Marmara University’s BYS system.**

------------------------------------------------------------------------

### **Academic Integrity & Ethical Guidelines**

To ensure a **fair, transparent, and ethical learning environment**, students must adhere to the following principles:

1.  **Academic Honesty** – Plagiarism and cheating are serious offenses and will not be tolerated. Any violations may result in disciplinary action.

2.  **Exam Conduct** – Students must follow all exam rules and refrain from using unauthorized materials (e.g., books, notes, electronic devices).

3.  **Original Work** – All assignments and exams must reflect the student's own work. Proper citation is required for any external sources.

4.  **Respect for Peers** – Students should engage respectfully with classmates and avoid unauthorized collaboration on assignments.

5.  **Instructor Communication** – If you have concerns about academic integrity or ethical dilemmas, discuss them openly with the instructor.

6.  **Accountability** – Students are responsible for their own learning, participation, and adherence to ethical guidelines.

7.  **Consequences of Violations** – Breaching these principles may result in academic penalties, such as assignment failure or disciplinary actions by the university.

Maintaining **integrity and fairness** in academics fosters a positive learning environment and supports students' intellectual growth.

------------------------------------------------------------------------

### **Accessing Google Classroom**

To join the course on **Google Classroom**, follow these steps:

1.  **Sign in to your Google Account** and use the invitation link:\
    👉 [Join Google Classroom](https://classroom.google.com/c/NTkxOTY2NDEzMjY4?cjc=3u3zpnd)

2.  Alternatively, you can manually join:

    -   Go to [Google Classroom](https://classroom.google.com).

    -   Sign in with your **Google Account**.

    -   On the "My Classes" page, click **"Join"**.

    -   Enter the **class code:** `xweaxw6` and click **"Join"**.

------------------------------------------------------------------------

### **Important Notes:**

-   Ensure that you enter the **correct class code** to avoid errors.

-   Once you join, you will have access to **assignments, announcements, and course materials**.

-   You can also access Google Classroom from your **mobile device**.

-   Make sure you are using the **latest version** of the Google Classroom app for a smooth experience.

------------------------------------------------------------------------

```{r setup, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(readr)
library(knitr)
# Load course schedule CSV
course_schedule <- read_csv("~/Desktop/mac_projects 2/data-science-with-R/Course_Schedule.csv")
```

```{r}
# Display the schedule as a formatted table
kable(course_schedule)
```

------------------------------------------------------------------------

### **Additional Information**

-   The course **content and structure** may be modified at the instructor’s discretion. Any changes will be communicated in advance.

-   Students are encouraged to reach out to the **instructor or teaching assistant** if they have any questions or concerns regarding the course. Contact details are provided above.

-   **Note:** This syllabus outlines the **general structure and teaching methods** of the course. Additional course materials and a detailed schedule will be provided as the course progresses.

**Wishing you a successful semester!** 🚀
